Revisiting the hydrodynamic modulation of short surface waves by longer waves

### 2025-03-05 paper.tex
Lagrangian gravity

\begin{table}
\begin{center}
\def~{\hphantom{0}}
\begin{tabular}{cccccc}
Reference   & Analysis   & Gravity & Nonlinear  & Nonlinear   & Solution \\
            & framework & varies  & long waves & short waves &          \\
\hline
\citet{longuet1960changes} & Velocity potential  & No  & No  & No  & Analytical \\
\hline
\citet{phillips1981dispersion} & Crest and action & Yes & Yes & No  & Numerical \\
                               & conservation     &     &     &     &           \\
\hline
\citet{longuet1987propagation} & Velocity potential & Yes & Yes & No  & Numerical \\
\hline
\citet{henyey1988energy} & Hamiltonian & Yes & Yes & No  & Analytical \\
\hline
\citet{zhang1990evolution} & Schr\" odinger & Yes & Yes & Yes & Numerical \\
\hline
\citet{peureux2021unsteady} & Crest and action & Yes & No  & No  & Numerical \\
                            & conservation     &     &     &     &           \\
\hline
This paper  & Crest and action & Yes & Yes & No  & Analytical \\
            & conservation     &     &     &     & and numerical \\
\hline
\end{tabular}
\caption{
  Non-exhaustive summary of prior and present (this paper) approaches to
  calculating the modulation of short waves by long waves.
}
\label{table:modulation_literature_summary}
\end{center}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/fig_analytical_solutions_ak0.1.pdf}
\caption{
  Steady solutions for the modulation of short-wave (a) wavenumber, (b)
  gravitational acceleration, (c) amplitude, (d) steepness, (e) intrinsic
  frequency, and (f) intrinsic phase speed as function of long-wave phase for
  $\varepsilon_L = 0.1$, based on
  \citet{longuet1960changes} (L-HS 1960, black) and this paper (blue).
  Long-wave crest and trough are located at $\psi = 0$ and $\psi = \pi$,
  respectively.
}
\label{fig:analytical_solutions_ak0.1}
\end{figure}

@@

Long waves induce orbital (centripetal) accelerations on their surface, so the
short waves ride in an accelerated reference frame and experience effective
gravity that is lower at the crests and higher in the troughs
\citep{longuet1986eulerian,longuet1987propagation}.
Further, the projection of the long-wave induced horizontal acceleration
contributes to small but non-negligible effects on the short-wave effective
gravity in the curvilinear coordinate system
\citep{phillips1981dispersion,zhang1990evolution}.
These two effects are completely described by projecting the gravitational
acceleration vector from the coordinate that is perpendicular to the mean water
surface ($z=0$) to that which is perpendicular to the long-wave surface
($z=\eta_L$), and likewise for the long-wave orbital accelerations:

\begin{equation}
\label{eq:gravity_modulation_general}
\frac{\widetilde{g}}{g}
  = \cos{\alpha} 
  + \dfrac{\partial W_{z=\eta}}{\partial t} \cos{\alpha}
  + \dfrac{\partial U_{z=\eta}}{\partial t} \sin{\alpha}
\end{equation}

\begin{equation}
\label{eq:local_slope}
\alpha = \tan^{-1}{\dfrac{\partial \eta}{\partial x}}
\end{equation}
where $\alpha$ is the local slope of the long-wave surface.
For short waves on a linear long wave, the gravity modulation is:

\begin{equation}
\label{eq:gravity_modulation_linear}
\frac{\widetilde{g}}{g}
  = 
  \frac{
    1 - \varepsilon_L \cos{\psi} e^{\varepsilon_L \cos{\psi}}
    \left[ 1 + \left(\varepsilon_L \sin{\psi}\right)^2 \right]
  }
  {\sqrt{\left(\varepsilon_L \sin{\psi}\right)^2 + 1}}
\end{equation}
Curvilinear effects on effective gravity can be removed altogether by setting
$\alpha = 0$.
In that case, (\ref{eq:gravity_modulation_general}) simplifies to:

@@

\begin{equation}
\label{eq:gravity_modulation_general_stokes}
\frac{\widetilde{g}}{g}
  = \cos{\alpha_{St}}
  + \dfrac{\partial W_{z=\eta_{St}}}{\partial t} \cos{\alpha_{St}}
  + \dfrac{\partial U_{z=\eta_{St}}}{\partial t} \sin{\alpha_{St}}
\end{equation}

@@

Let us now examine the contributions of the different terms to the short-wave
effective gravity modulation.
Fig. \ref{fig:effective_gravities}a shows the effective gravity modulation of
short-wave as a function of the long-wave phase for $\varepsilon_L = 0.2$, for
a variety of the gravity modulation forms.
This value of $\varepsilon_L$ allows easier visualisation of the modulation
differences, while still being relatively moderate.
To the first order in $\varepsilon_L$, the gravity modulation is out of phase
with the long-wave elevation, with the short waves experiencing a lower effective
gravity on the long-wave crests and a higher effective gravity in the troughs.
Evaluating the orbital accelerations at $z=\eta_L$ causes a gravity reduction
around the crests ($\approx6\%$) and the troughs ($\approx 3\%$), and an
increase on the front and rear faces ($\approx 4\%$) of the long wave, due to
the exponential change in the orbital accelerations away from the mean water level
(Fig. \ref{fig:effective_gravities}b).
The difference between the linear and Stokes long waves at this steepness is
negligible.
The long-wave elevation curvature has a second-order effect in $\varepsilon_L$.
Naturally, there is no curvature exactly at the crests and the troughs, and its
effects are largest at the front and rear faces of the long wave, where the
covariance between the local slope and the horizontal accelerations are largest.

@@

It is instructive to also examine the mean, maximum, and minimum values of the
effective gravity as function of long-wave steepness $\varepsilon_L$
(Fig. \ref{fig:effective_gravities_mean_max_min}).
The long-wave averaged effective gravity of short waves remains equal to the
resting value of $g$ for all values of $\varepsilon_L$, except if the effects
of curvature are considered, in which case the average effective gravity is
reduced.
This reduction, in case of a long Stokes wave, is $\approx 3\%$ for
$\varepsilon_L = 0.2$ and $\approx 14\%$ for $\varepsilon_L = 0.4$.
As the effective gravity minima and maxima occur at crests and troughs,
respectively, the curvature affects the phase-averaged values but not the extremes.
As the slope effects on the effective gravity exist only away from crests and
troughs, they may be neglected in the steady solutions where modulation is
typically evaluated at the long-wave crests. 
The effective gravity reduction at the crests is $\approx 11\%$, $25\%$, and $66\%$
for $\varepsilon_L = 0.1$, $0.2$, and $0.4$, respectively.
These are similar to the fully-nonlinear numerical estimates of
\citet{longuet1986eulerian,longuet1987propagation}.
Note that these are strictly Eulerian values, evaluated at fixed positions in
space.
As \citet{longuet1986eulerian} pointed out, the effective gravity in the
Lagrangian frame (\textit{i.e.}, following the surface particles) is smoother
and less amplified because the moving particles spend a longer time riding near
the crests and less time returning through the troughs.
However, when we solve the short-wave crest and action conservation equations
in an Eulerian frame, it is the Eulerian effective gravity that must be applied
to conserve the short-wave properties across the long-wave phase.

@@

Following the variational approach by \citet{whitham1965general},
\citet{bretherton1968wavetrains} showed that the wave action is conserved for
small-amplitude (linear) waves that propagate in slowly-varying media.
\citet{longuet1987propagation} correctly pointed out both
requirements--that both the medium and the short wavetrain must be
slowly-varying--for the wave action conservation to hold.
However, he incorrectly stated that there is no explicit restriction on the
steepness of the long wave, and that it appears necessary to only assume
small-amplitude short waves and significant scale separation ($k/k_L$).
The restriction on $\varepsilon_L$ becomes readily apparent when we recognize
that the long wave-induced orbital velocity scales with $\varepsilon_L$, and
that the divergence of this velocity is the dominant term in (\ref{eq:wavenumber})
and (\ref{eq:wave_action2}).
Here we show that in the general case both the homogeneity and the stationarity
of a short wave quantity $q$ depend on the long-wave steepness $\varepsilon_L$.

@@

\backsection[Funding]{
  Milan Curcic was partly supported by the National Science Foundation
  grant AGS-1745384, Transatlantic Research Partnership by the French Embassy
  in the United States and the FACE Foundation, and the Office of Naval Research
  Coastal Land Air-Sea Interaction DRI.
}

@@

The long-wave induced velocities (\ref{eq:phi})-(\ref{eq:W_L}) are accurate to
$\mathcal{O}(\varepsilon_L^3)$ when evaluated at the mean water level.
However, in this paper we depart from the mean water level and evaluate the
velocities at the surface of the long wave.
This introduces a additional errors at $\mathcal{O}(\varepsilon_L^2)$.
To quantify this error exactly, we compare the velocities evaluated at the
surface for a linear (small-amplitude) wave and a third-order Stokes wave,
with the surface velocity of a fully nonlinear wave
(Fig. \ref{fig:velocity_error_by_phase}).
The elevation and the surface velocity for the fully nonlinear wave is
calculated numerically using the method by \citet{clamond2018accurate},
which relies on Petviashvili's iterations and in this case uses 2048 Fourier
modes to represent the wave.
Both the 1$^{st}$ and the 3$^{rd}$ order waves underestimate the surface velocity
at the crest and in the trough, and overestimate it on the front and rear faces
of the crest.

>>>>

\begin{table}
\begin{center}
\def~{\hphantom{0}}
\begin{tabular}{cccccc}
Reference   & Analysis   & Gravity & Nonlinear  & Nonlinear   & Solution \\
            & framework & varies  & long waves & short waves &          \\
\hline
\citet{longuet1960changes} & Velocity potential  & No  & No  & No  & Analytical \\
\hline
\citet{phillips1981dispersion} & Crest and action & Yes & Yes & No  & Numerical \\
                               & conservation     &     &     &     &           \\
\hline
\citet{longuet1987propagation} & Crest and action & Yes & Yes & No  & Numerical \\
                               & conservation     &     &     &     &           \\
\hline
\citet{henyey1988energy} & Hamiltonian & Yes & Yes & No  & Analytical \\
\hline
\citet{zhang1990evolution} & Schr\" odinger & Yes & Yes & Yes & Numerical \\
\hline
\citet{peureux2021unsteady} & Crest and action & Yes & No  & No  & Numerical \\
                            & conservation     &     &     &     &           \\
\hline
This paper  & Crest and action & Yes & Yes & No  & Analytical \\
            & conservation     &     &     &     & and numerical \\
\hline
\end{tabular}
\caption{
  Non-exhaustive summary of prior and present (this paper) approaches to
  calculating the modulation of short waves by long waves.
}
\label{table:modulation_literature_summary}
\end{center}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/fig_analytical_solutions_ak0.1.pdf}
\caption{
  Steady analytical solutions for the modulation of short-wave (a) wavenumber, (b)
  gravitational acceleration, (c) amplitude, (d) steepness, (e) intrinsic
  frequency, and (f) intrinsic phase speed as function of long-wave phase for
  $\varepsilon_L = 0.1$, based on
  \citet{longuet1960changes} (L-HS 1960, black) and this paper (blue).
  Long-wave crest and trough are located at $\psi = 0$ and $\psi = \pi$,
  respectively.
}
\label{fig:analytical_solutions_ak0.1}
\end{figure}

@@

Three distinct effects contribute to the modulation of the effective gravity of
short waves in the presence of long waves.
First, long waves induce orbital (centripetal) accelerations on their surface,
so the short waves ride in an accelerated reference frame and experience effective
gravity that is lower at the crests and higher in the troughs
\citep{longuet1986eulerian,longuet1987propagation}.
This is an $\mathcal{O}(\varepsilon_L)$ effect.
Second, it is important to distinguish between the Eulerian and Lagrangian effective
gravities \citep{longuet1986eulerian}.
The Eulerian gravity is that which would be measured by a fixed wave staff,
for example, and features sharp and strong minima at the crests and broad
and weak maxima in the troughs.
In contrast, the Lagrangian gravity is that which is experienced by a short wave
group that propagates with its own group speed and is advected by the ambient,
long-wave induced, velocity.
The Lagrangian velocity is overall smoother and more attenuated compared to
its Eulerian counterpart because the local slope as experienced by the
travelling short-wave group is gentler.
The two gravities are related by:

\begin{equation}
\label{eq:gravity_modulation_lagrangian}
\widetilde{g_l} =
g + \frac{dW}{dt} =
g + \frac{\partial W}{\partial t} + \left(C_g + U\right) \frac{\partial W}{\partial x} =
\widetilde{g_e} + \left(C_g + U\right) \frac{\partial W}{\partial x}
\end{equation}
where $l$ and $e$ subscripts serve to denote "Lagrangian" and "Eulerian",
respectively.
Although \citet{longuet1986eulerian} thoroughly described the differences
between the Eulerian and Lagrangian effective gravities, he did so for
a fully nonlinear irrotational gravity wave, which can be computed numerically
but is not readily available analytically.
Note also that when evaluating the Lagrangian effective gravity,
\citet{longuet1987propagation} considered the orbital motion of the long wave
but not the propagation speed of the short-wave group, which can exceed the
ambient velocity $U$ depending on the properties of the short and long waves
considered.
Group speed of the short waves aside, the Lagrangian correction to the gravity
being an advective term is $\mathcal{O}(\varepsilon_L^2)$.
Third, the projection of the long-wave induced horizontal acceleration
contributes small but non-negligible effects on the short-wave effective
gravity in the curvilinear coordinate system
\citep{phillips1981dispersion,zhang1990evolution}.
These three effects are completely described by projecting the gravitational
acceleration vector from the coordinate that is perpendicular to the mean water
surface ($z=0$) to that which is perpendicular to the long-wave surface
($z=\eta_L$), and likewise for the long-wave orbital accelerations:

\begin{equation}
\label{eq:gravity_modulation_general}
\frac{\widetilde{g}}{g}
  = \cos{\alpha} 
  + \frac{1}{g} \dfrac{dW_{z=\eta_L}}{dt} \cos{\alpha}
  + \frac{1}{g} \dfrac{dU_{z=\eta_L}}{dt} \sin{\alpha}
\end{equation}

\begin{equation}
\label{eq:local_slope}
\alpha = \tan^{-1}{\dfrac{\partial \eta}{\partial x}}
\end{equation}
where $\alpha$ is the local slope of the long-wave surface.
For short waves on a linear long wave, the Eulerian gravity modulation is:

\begin{equation}
\label{eq:gravity_modulation_linear}
\frac{\widetilde{g_e}}{g}
  = 
  \frac{
    1 - \varepsilon_L \cos{\psi} e^{\varepsilon_L \cos{\psi}}
    \left[ 1 + \left(\varepsilon_L \sin{\psi}\right)^2 \right]
  }
  {\sqrt{\left(\varepsilon_L \sin{\psi}\right)^2 + 1}}
\end{equation}
The same derivation can be carried out for the Lagrangian gravity
modulation following (\ref{eq:gravity_modulation_lagrangian})-(\ref{eq:local_slope}),
and is omitted here for brevity.
Curvilinear effects on effective gravity can be removed altogether by setting
$\alpha = 0$.
In that case, (\ref{eq:gravity_modulation_general}) simplifies to:

@@

\begin{equation}
\label{eq:gravity_modulation_general_stokes}
\frac{\widetilde{g}}{g}
  = \cos{\alpha_{St}}
  + \dfrac{dW_{z=\eta_{St}}}{dt} \cos{\alpha_{St}}
  + \dfrac{dU_{z=\eta_{St}}}{dt} \sin{\alpha_{St}}
\end{equation}

@@

Let us now examine the contributions of the different terms to the short-wave
effective gravity modulation.
Fig. \ref{fig:effective_gravities}a shows the effective gravity modulation of
short waves as a function of the long-wave phase for $\varepsilon_L = 0.2$, for
a variety of the gravity modulation forms.
This value of $\varepsilon_L$ allows easier visualisation of the modulation
differences, while still being relatively moderate.
To the first order in $\varepsilon_L$, the gravity modulation is out of phase
with the long-wave elevation, with the short waves experiencing a lower effective
gravity on the long-wave crests and a higher effective gravity in the troughs.
Evaluating the orbital accelerations at $z=\eta_L$ causes a gravity reduction
around the crests ($\approx6\%$) and the troughs ($\approx 3\%$), and an
increase on the front and rear faces ($\approx 4\%$) of the long wave, due to
the exponential change in the orbital accelerations away from the mean water level
(Fig. \ref{fig:effective_gravities}b).
The difference between the linear and Stokes long waves at this steepness is
negligible.
The long-wave elevation curvature has a second-order effect in $\varepsilon_L$.
Naturally, there is no curvature exactly at the crests and the troughs, and its
effects are largest at the front and rear faces of the long wave, where the
covariance between the local slope and the horizontal accelerations are largest.

@@

It is instructive to also examine the mean, maximum, and minimum values of the
effective gravity as function of long-wave steepness $\varepsilon_L$
(Fig. \ref{fig:effective_gravities_mean_max_min}).
The long-wave averaged effective gravity of short waves remains equal to the
resting value of $g$ for all values of $\varepsilon_L$, except if the effects
of curvature are considered, in which case the average effective gravity is
reduced.
This reduction, in case of a long Stokes wave, is $\approx 3\%$ for
$\varepsilon_L = 0.2$ and $\approx 14\%$ for $\varepsilon_L = 0.4$.
As the effective gravity minima and maxima occur at crests and troughs,
respectively, the curvature affects the phase-averaged values but not the extremes.
As the slope effects on the effective gravity exist only away from crests and
troughs, they may be neglected in the steady solutions where modulation is
typically evaluated at the long-wave crests. 
The effective gravity reduction at the crests is $\approx 11\%$, $25\%$, and $66\%$
for $\varepsilon_L = 0.1$, $0.2$, and $0.4$, respectively.
These are similar to the fully-nonlinear numerical estimates of
\citet{longuet1986eulerian,longuet1987propagation}.

@@

Following the variational approach by \citet{whitham1965general},
\citet{bretherton1968wavetrains} showed that the wave action is conserved for
small-amplitude (linear) waves that propagate in slowly-varying media.
\citet{longuet1987propagation} correctly pointed out both
requirements--that both the medium and the short wavetrain must be
slowly-varying--for the wave action conservation to hold.
However, he incorrectly stated that there is no explicit restriction on the
steepness of the long wave, and that it appears necessary to only assume
small-amplitude short waves and significant scale separation ($k/k_L$).
The restriction on $\varepsilon_L$ becomes apparent when we recognize
that the long wave-induced orbital velocity scales with $\varepsilon_L$, and
that the divergence of this velocity is the dominant term in (\ref{eq:wavenumber})
and (\ref{eq:wave_action2}).
Here we show that in the general case both the homogeneity and the stationarity
of a short wave quantity $q$ depend on the long-wave steepness $\varepsilon_L$.

@@

\backsection[Funding]{
  Milan Curcic was partly supported by the National Science Foundation
  grant AGS-1745384, Transatlantic Research Partnership by the French Embassy
  in the United States and the FACE Foundation, and the Office of Naval Research
  grants N000142012102 (Coastal Land Air-Sea Interaction DRI) and
  N000142412598 (in collaboration with SASCWATCH: Study on Air-Sea Coupling with
  WAves, Turbulence, and Clouds at High winds MURI).
}

@@

The long-wave induced velocities (\ref{eq:phi})-(\ref{eq:W_L}) are accurate to
$\mathcal{O}(\varepsilon_L^3)$ when evaluated at the mean water level.
However, in this paper we depart from the mean water level and evaluate the
velocities at the surface of the long wave.
This introduces additional errors at $\mathcal{O}(\varepsilon_L^2)$.
To quantify this error exactly, we compare the velocities evaluated at the
surface for a linear (small-amplitude) wave and a third-order Stokes wave,
with the surface velocity of a fully nonlinear wave
(Fig. \ref{fig:velocity_error_by_phase}).
The elevation and the surface velocity for the fully nonlinear wave is
calculated numerically using the method by \citet{clamond2018accurate},
which relies on Petviashvili's iterations and in this case uses 2048 Fourier
modes to represent the wave.
Both the 1$^{st}$ and the 3$^{rd}$ order waves underestimate the surface velocity
at the crest and in the trough, and overestimate it on the front and rear faces
of the crest.
